\stktokens{}: Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities

### 2018-10-15 fully-abstract-comp/popl2019/paper.tex
Some small changes.

Local state encapsulation (LSE) is the guarantee that when a function invokes another function, its local variables (saved on its stack frame) will not have been read or modified when the invoked function returns.
At the assembly level, this property also does not hold immediately.
The calling function's local variables are stored on the stack during the invocation, and functions are not supposed to touch stack frames other than their own.
However, untrusted assembly code is free to ignore this requirement and read or overwrite the local state of other stack frames.
To enforce these properties, target language security primitives are needed that can be used to prevent untrusted code from misbehaving without imposing too much overhead on well-behaved code.
The virtual-memory based security primitives on commodity processors do not seem sufficiently fine-grained to efficiently support this.
More suitable security primitives are offered by a type of CPUs known as capability machines \citep{levy_capability-based_1984,watson_cheri:_2015}.
These processors use tagged memory to enforce a strict separation between integers and {\itshape capabilities}: pointers that carry authority.
Capabilities come in different flavours.
Memory capabilities allow reading from and writing to a block of memory.
Additionally, capability machines offer some form of {\itshape object capabilities} that represent low-level encapsulated closures, i.e. a piece of code coupled with private state that it gains access to upon invocation.
The concrete mechanics of object capabilities vary between different capability machines.
% KJAA: This 'sealed' and 'common seal' business is unclear to me.
For example, on a recent capability machine called CHERI they take the form of pairs of capabilities that represent the code and data parts of the closure.
Each of the two capabilities are sealed with a common seal which make them opaque.
The hardware transparently unseals the pair upon invocation~\citep{watson_capability_2015}.

@@

A more scalable approach retains a single stack shared between components.
Enforcing WBCF and LSE in this approach requires a way to temporarily provide stack and return capabilities to an untrusted component and to revoke them after it returns.
While capability revocation is expensive in general, some capability machines offer restricted forms of revocation that can be implemented efficiently.
For example, CHERI offers a form of {\itshape local} capabilities that can only be stored in registers or on the stack but not in other parts of memory.
\citet{skorstengaard_reasoning_2017} has demonstrated that by making the stack and return pointer local, and by introducing a number of security checks and measures, the two properties can be guaranteed.
In fact, Robert Watson told us privately that a similar system was envisioned in earlier work on CHERI~\citep{watson2012cheri}. 
% KJAA: 'Boundary crossing' is new here.
However, a problem with this approach is that revoking the local stack and return capabilities on every security boundary crossing requires clearing the entire unused part of the stack, an operation that may be prohibitively expensive.

In this work, we propose and study \stktokens{}: an alternative calling convention that enforces WBCF and LSE with a single shared stack.
Instead of CHERI's local capabilities, it builds on {\itshape linear} capabilities; a new form of capabilities that has not been previously described in the published literature.\footnote{Although they have been mentioned in some technical documents.}
% KJAA: Perhaps make it clear by who and when/where?
The hardware prevents these capabilities from being duplicated.
We propose to make stack and return pointers linear and require components to hand them out in cross-component invocations and to return them in returns.
The non-duplicability of linear capabilities together with some security checks allow us to guarantee WBCF and LSE without large overhead on boundary crossings in particular without the need for clearing large blocks of memory.

@@

%The main contributions of this paper are (1) \stktokens{} a calling convention that uses linear capabilities to provably secure LSE and WBCF, and (2) a novel proof technique that we call fully-abstract overlay semantics which we apply to prove the properties of \stktokens{}.
In this paper, we make the following contributions:
% Presents the first formal model of a capability machine with linear capabilities (Section 2)
% Presents a calling convention called \stktokens{} that provably guarantee local state encapsulation and well-bracketed control flow (Section 3)
% Presents a novel proof technique that we call fully-abstract overlay semantics. (Section 4-5)
\begin{itemize}
\item We present \trgcm{}: A formalization of a simple CHERI-like capability machine with linear capabilities (Section~\ref{sec:cap-mach-w-seal-and-lin}).
\item We present a new calling convention \stktokens{} that provably guarantees LSE and WBCF on \trgcm{} (Section~\ref{sec:stktokens-explained}).
\item We present a new way to formalize these guarantees based on a novel
  technique called \textit{fully-abstract overlay semantics} and we prove LSE
  and WBCF claims. This includes:
  \begin{itemize}
  \item defining \srccm{}: an overlay semantics for \trgcm{} with built-in LSE and WBCF (Section~\ref{sec:form-secur-with})
  \item proving full-abstraction for a compiler from \srccm{} to \trgcm{} (Section~\ref{sec:fa-proof}) by
  \item using and defining a cross-language, step-indexed, Kripke logical relation with recursive worlds (Section~\ref{sec:fa-proof}).
  \end{itemize}
\end{itemize}
This paper is accompanied by a technical report~\citep{technical_report} with the elided details and proof.

@@

\section{A Capability Machine with Sealing and Linear Capabilities}
\label{sec:cap-mach-w-seal-and-lin}
% Brief intro to section and this cap machine:
% Cap machine inspired by CHERI, but with linear capabilities
\begin{wrapfigure}{r}{0.26\linewidth}
  \centering
  \begin{tikzpicture}[main node/.style={}]
    \node[main node] (rwx) {$\rwx$};
    \node[main node] (rx) [below right of=rwx] {$\rx$};

@@

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{wrapfigure}
In this section, we introduce a simple but representative capability machine with linear capabilities, that we call \trgcm{} (Linear Capability Machine).
\trgcm{} is mainly inspired by CHERI~\citep{watson_cheri:_2015} with linear capabilities as the main addition.
For simplicity, \trgcm{} assumes an infinite address space and unbounded integers.
% \lau{Do we have a good reason to assume an infinite address space for this work?
%   We do not model malloc, so we do not need to worry about what happens when malloc runs out of memory.}

%%% Capabilities
The concept of a capability is the cornerstone of any capability machine.
In its simplest form, a capability is a permission and a range of authority.
The permission dictates the operations the capability can be used for, and the range of authority specifies the range of memory it can act upon.
% Permissions and Range of authority
The capabilities on \trgcm{} are of the form $((\var{perm},\var{lin}),\var{base},\var{end},\var{addr})$ (defined in Figure~\ref{fig:target-syntax} with the rest of the syntax of \trgcm{}). Here $\var{perm}$ is the permission, and $[\var{base},\var{end}]$ is the range of authority.
The available permissions are read-write-execute ($\rwx$), read-write ($\rw$), read-execute ($\rx$), read-only ($\ro$), and null-permission ($\noperm$) ordered by $\le$ as illustrated in Figure~\ref{fig:perm-hier}.
% Pointer and linearity
In addition to the permission and range, capabilities also have a current address $\var{addr}$ and a linearity $\var{lin}$.
The linearity is either $\normal$ for traditional capabilities or $\linear$ for linear ones.
% What is linear capabilities, split/splice
A linear capability is a capability that cannot be duplicated.
This is enforced dynamically on the capability machine, so when a linear capability is moved between registers or memory, the source is cleared.
The non-duplicability of linear capabilities means that a linear capability cannot become aliased if it wasn't to begin with.

>>>>

Local state encapsulation (LSE) is the guarantee that when a function invokes another function, its local variables (saved on its stack frame) will not have been read or modified when the invoked function returns.
At the assembly level, this property also does not hold immediately.
The calling function's local variables are stored on the stack during the invocation, and functions are not supposed to touch stack frames other than their own.
However, untrusted assembly code is free to ignore this requirement and read or overwrite the local state of other stack frames.
To enforce these properties, target language security primitives are needed that can be used to prevent untrusted code from misbehaving without imposing too much overhead on well-behaved code.
The virtual-memory based security primitives on commodity processors do not seem sufficiently fine-grained to efficiently support this.
More suitable security primitives are offered by a type of CPUs known as capability machines \citep{levy_capability-based_1984,watson_cheri:_2015}.
These processors use tagged memory to enforce a strict separation between integers and {\itshape capabilities}: pointers that carry authority.
Capabilities come in different flavours.
Memory capabilities allow reading from and writing to a block of memory.
Additionally, capability machines offer some form of {\itshape object capabilities} that represent low-level encapsulated closures, i.e. a piece of code coupled with private state that it gains access to upon invocation.
The concrete mechanics of object capabilities vary between different capability machines.
% KJAA: This 'sealed' and 'common seal' business is unclear to me.
For example, on a recent capability machine called CHERI they take the form of pairs of capabilities that represent the code and data parts of the closure.
Each of the two capabilities are sealed with a common seal which make them opaque.
The hardware transparently unseals the pair upon invocation~\citep{watson_capability_2015,watson_fast_2016}.

@@

A more scalable approach retains a single stack shared between components.
Enforcing WBCF and LSE in this approach requires a way to temporarily provide stack and return capabilities to an untrusted component and to revoke them after it returns.
While capability revocation is expensive in general, some capability machines offer restricted forms of revocation that can be implemented efficiently.
For example, CHERI offers a form of {\itshape local} capabilities that can only be stored in registers or on the stack but not in other parts of memory.
\citet{skorstengaard_reasoning_2017} have demonstrated that by making the stack and return pointer local, and by introducing a number of security checks and measures, the two properties can be guaranteed.
In fact, a similar system was envisioned in earlier work on CHERI~\citep{watson2012cheri}. 
% KJAA: 'Boundary crossing' is new here.
However, a problem with this approach is that revoking the local stack and return capabilities on every security boundary crossing requires clearing the entire unused part of the stack, an operation that may be prohibitively expensive.

In this work, we propose and study \stktokens{}: an alternative calling convention that enforces WBCF and LSE with a single shared stack.
Instead of CHERI's local capabilities, it builds on {\itshape linear} capabilities; a new form of capabilities that has not been previously described in the published literature.\footnote{Although they have been mentioned in some technical documents.}
% KJAA: Perhaps make it clear by who and when/where?
The hardware prevents these capabilities from being duplicated.
We propose to make stack and return pointers linear and require components to hand them out in cross-component invocations and to return them in returns.
The non-duplicability of linear capabilities together with some security checks allow us to guarantee WBCF and LSE without large overhead on boundary crossings and in particular without the need for clearing large blocks of memory.

@@

%The main contributions of this paper are (1) \stktokens{} a calling convention that uses linear capabilities to provably secure LSE and WBCF, and (2) a novel proof technique that we call fully-abstract overlay semantics which we apply to prove the properties of \stktokens{}.
In this paper, we make the following contributions:
% Presents the first formal model of a capability machine with linear capabilities (Section 2)
% Presents a calling convention called \stktokens{} that provably guarantee local state encapsulation and well-bracketed control flow (Section 3)
% Presents a novel proof technique that we call fully-abstract overlay semantics. (Section 4-5)
\begin{itemize}
\item We present \trgcm{}: A formalization of a simple CHERI-like capability machine with linear capabilities (Section~\ref{sec:cap-mach-w-seal-and-lin}).
\item We present a new calling convention \stktokens{} that provably guarantees LSE and WBCF on \trgcm{} (Section~\ref{sec:stktokens-explained}).
\item We present a new way to formalize these guarantees based on a novel
  technique called \textit{fully-abstract overlay semantics} and we prove LSE
  and WBCF claims. This includes:
  \begin{itemize}
  \item defining \srccm{}: an overlay semantics for \trgcm{} with built-in LSE and WBCF (Section~\ref{sec:form-secur-with})
  \item proving full-abstraction for the embedding of \srccm{} into \trgcm{} (Section~\ref{sec:fa-proof}) by
  \item using and defining a cross-language, step-indexed, Kripke logical relation with recursive worlds (Section~\ref{sec:fa-proof}).
  \end{itemize}
\end{itemize}
This paper is accompanied by a technical report~\citep{technical_report} with the elided details and proof.

@@

\section{A Capability Machine with Sealing and Linear Capabilities}
\label{sec:cap-mach-w-seal-and-lin}
% Brief intro to section and this cap machine:
% Cap machine inspired by CHERI, but with linear capabilities
In this section, we introduce a simple but representative capability machine with linear capabilities, that we call \trgcm{} (Linear Capability Machine).
\trgcm{} is mainly inspired by CHERI~\citep{watson_cheri:_2015} with linear capabilities as the main addition.
For simplicity, \trgcm{} assumes an infinite address space and unbounded integers.
% \lau{Do we have a good reason to assume an infinite address space for this work?
%   We do not model malloc, so we do not need to worry about what happens when malloc runs out of memory.}

%%% Capabilities
\begin{wrapfigure}{r}{0.26\linewidth}
  \centering
  \begin{tikzpicture}[main node/.style={}]
    \node[main node] (rwx) {$\rwx$};
    \node[main node] (rx) [below right of=rwx] {$\rx$};

@@

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{wrapfigure}
The concept of a capability is the cornerstone of any capability machine.
In its simplest form, a capability is a permission and a range of authority.
The permission dictates the operations the capability can be used for, and the range of authority specifies the range of memory it can act upon.
% Permissions and Range of authority
The capabilities on \trgcm{} are of the form $((\var{perm},\var{lin}),\var{base},\var{end},\var{addr})$ (defined in Figure~\ref{fig:target-syntax} with the rest of the syntax of \trgcm{}). Here $\var{perm}$ is the permission, and $[\var{base},\var{end}]$ is the range of authority.
The available permissions are read-write-execute ($\rwx$), read-write ($\rw$), read-execute ($\rx$), read-only ($\ro$), and null-permission ($\noperm$) ordered by $\le$ as illustrated in Figure~\ref{fig:perm-hier}.
% Pointer and linearity
In addition to the permission and range, capabilities also have a current address $\var{addr}$ and a linearity $\var{lin}$.
The linearity is either $\normal$ for traditional capabilities or $\linear$ for linear ones.
% What is linear capabilities, split/splice
A linear capability is a capability that cannot be duplicated.
This is enforced dynamically on the capability machine, so when a linear capability is moved between registers or memory, the source is cleared.
The non-duplicability of linear capabilities means that a linear capability cannot become aliased if it wasn't to begin with.
